---
title: "MATH532 Practical 2 (Student) — Epidemics on Networks"
output:
html_document:
toc: true
toc_depth: 3
number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Setup

```{r libraries}
suppressPackageStartupMessages({
  library(igraph)
  library(dplyr)
  library(tibble)
  library(ggplot2)
})

set.seed(124)

# Put Network_model.R in the same folder as this Rmd
source("Network_model.R")
```

```{r helpers}
# Small helpers for plotting and summaries
estimate_r_early <- function(df, t_max = 8) {
  # TODO (Section B):
  # Estimate an early exponential growth rate r by fitting:
  #   log(prevalence) ~ time
  # on a short early window (time <= t_max)
  #
  # Hints:
  # - filter to time <= t_max
  # - use y <- pmax(prevalence, 1e-9) to avoid log(0)
  # - fit <- lm(log(y) ~ time, data = ...)
  # - return slope coefficient
  stop("TODO: implement estimate_r_early(df, t_max)")
}

gc_fraction <- function(A) {
  # GC fraction for an adjacency matrix A
  g <- graph_from_adjacency_matrix(A, mode = "undirected", diag = FALSE)
  max(components(g)$csize) / nrow(A)
}
```

---

## Section A — Epidemics across graph families (guided)

### Instructions

1. Construct 4 graphs (ER, small-world, preferential attachment, modular).
2. Simulate SIR outbreaks with **neutral calibration**:

   * `calib_mode = "pre_removal_biology"`
3. Plot prevalence trajectories by graph family.
4. Write 2–3 bullets comparing shapes (peak timing, peak height, persistence).

### Provided code (run as-is)

```{r A-graphs}
N <- 500

A_er  <- Matrix(makeNetwork("rand", N=N), sparse=TRUE)

# Small-world graph with a custom rewiring probability
g_ws <- sample_smallworld(1, N, nei = 5, p = 0.02)
g_ws <- as.undirected(g_ws, mode = "collapse")
g_ws <- simplify(g_ws)
A_ws <- as_adj(g_ws, sparse = TRUE)

A_pa  <- Matrix(makeNetwork("prefAtt", N=N), sparse=TRUE)
A_mod <- Matrix(makeNetwork("randMod", N=N), sparse=TRUE)

graphs <- list(ER = A_er, WS = A_ws, PA = A_pa, Modular = A_mod)

TSPAN <- seq(0, 40, by = 0.1)
```

```{r A-run}
res_list <- lapply(names(graphs), function(nm) {
  A <- graphs[[nm]]
  out <- runSimulation(
    A, R0 = 2,
    frac = 0,
    removal_by = "rand",
    calib_mode = "pre_removal_biology",
    modelType = "SIR",
    T_R = 3, T_E = 2,
    TSPAN = TSPAN,
    n_runs = 3,
    seed_mode = "fixed",
    fixed_nodes = 1:10,
    n_seeds = 10,
    global_seed = 99
  )
  out$graph <- nm
  out
})

df_all <- bind_rows(res_list)

ggplot(df_all, aes(time, prevalence, color = sim)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~ graph, scales = "free_y") +
  theme_minimal() +
  labs(title="Epidemics across graph families (neutral calibration)",
       y="Prevalence", x="Time")
```

### Reflection A (2–3 bullets)

* Which graph produces the earliest peak? Which produces the latest?
* Which produces the highest peak? Any surprises?

---

## Section B — Structural proxy vs observed early growth (medium DIY)

### Instructions

1. Compute `neutral_moments()` for each graph.
2. Interpret:

   * `knbr - 1` as a **structural epidemic potential proxy** (neutral approximation).
3. Implement an early growth estimator `estimate_r_early()` and compare across graphs.

```{r B-moments}
mom_tbl <- lapply(names(graphs), function(nm) {
  A <- graphs[[nm]]
  m <- neutral_moments(A)
  tibble(graph = nm, mk = m$mk, vk = m$vk, knbr = m$knbr, proxy = (m$knbr - 1))
}) |> bind_rows()

mom_tbl
```

```{r B-growth, eval=FALSE}
# Runs after you implement estimate_r_early()

r_tbl <- df_all %>%
  group_by(graph, sim) %>%
  summarise(r_hat = estimate_r_early(cur_data_all(), t_max = 8), .groups = "drop") %>%
  group_by(graph) %>%
  summarise(r_median = median(r_hat, na.rm = TRUE), .groups = "drop")

r_tbl
```

### Reflection B

* Does the proxy (`knbr - 1`) agree with the ranking of early growth rates?
* Why might clustering or modularity weaken the proxy?

---

## Section C — Two calibration modes under degree-targeted removal (higher DIY)

### Instructions

We will compare two calibration logics:

* **Mode 1: `pre_removal_biology`**
  Calibrate (\beta) on the original network, then remove nodes and keep (\beta) fixed.
  (“Same pathogen; changed system.”)

* **Mode 2: `post_removal_sameR0`**
  Remove nodes first, then recalibrate (\beta) to force the same headline (R_0).
  (“Same (R_0) by construction; not the same pathogen.”)

**Task:** Implement the nested loop and plot prevalence vs time.

```{r C-stub, eval=FALSE}
# TODO (Section C): write a nested loop over frac_vals and mode_vals

A_base <- graphs[["PA"]]
frac_vals <- c(0.0, 0.1, 0.2, 0.3)
mode_vals <- c("pre_removal_biology","post_removal_sameR0")

# TODO: build df_modes with columns time, prevalence, sim, frac, mode
# Hints:
# - for each mode_ in mode_vals:
#   - for each fr in frac_vals:
#       out <- runSimulation(A_base, R0=2, frac=fr, removal_by="degree",
#                            calib_mode=mode_, modelType="SIR",
#                            T_R=3, TSPAN=TSPAN, n_runs=3,
#                            seed_mode="fixed", fixed_nodes=1:10,
#                            n_seeds=10, global_seed=123)
#       add columns frac and mode, then bind_rows

# df_modes <- ...

ggplot(df_modes, aes(time, prevalence, group = sim, color = factor(frac))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ mode, scales = "free_y") +
  theme_minimal() +
  labs(title="Calibration modes under degree-targeted removal",
       x="Time", y="Prevalence", color="Removed fraction")
```

### Reflection C

* Which calibration mode makes the intervention look “more effective”, and why?
* Which mode corresponds to “same pathogen biology”?

---

## Section D — Random vs degree-targeted removal + giant component (higher DIY)

### Instructions

1. For the PA graph, remove fractions (f\in{0,0.1,0.2,0.3}).
2. Compare `by="rand"` vs `by="degree"`.
3. Plot **giant component fraction** vs removed fraction.
4. Explain why this relates to epidemic size.

```{r D-stub, eval=FALSE}
# TODO (Section D): implement gc_tbl

A0 <- graphs[["PA"]]
removals <- expand.grid(
  frac = c(0.0, 0.1, 0.2, 0.3),
  by = c("rand","degree"),
  KEEP.OUT.ATTRS = FALSE
)

# gc_tbl <- ...

ggplot(gc_tbl, aes(frac, GCfrac, color = by)) +
  geom_line() + geom_point() +
  theme_minimal() +
  labs(title="Giant component fraction vs node removal",
       x="Removed fraction", y="GC fraction")
```

### Reflection D

* Why does degree-targeted removal reduce GC fraction faster on PA graphs?

---

## Section E — Ring lattice vs small-world: wavefront vs shortcuts (core demo)

### Instructions

1. Construct a **ring lattice** (WS with (p=0)) and a **small-world** graph (tiny (p)).
2. Run simulations with the same calibration mode (`pre_removal_biology`).
3. Compare growth shapes and comment on “linear-ish” wavefront growth vs shortcut seeding.

```{r E-stub, eval=FALSE}
# TODO (Section E): build A_ring and A_sw and run both

nei <- 5

# Ring lattice (p=0)
# g_ring <- sample_smallworld(1, N, nei=nei, p=0)
# ...

# Small-world with very small p
# g_sw <- sample_smallworld(1, N, nei=nei, p=0.01)
# ...

# df_sw <- bind_rows(
#   runSimulation(A_ring, R0=2, calib_mode="pre_removal_biology", T_R=3, TSPAN=TSPAN, n_runs=2,
#                 seed_mode="fixed", fixed_nodes=1:10, n_seeds=10, global_seed=7) |> mutate(kind="Ring (p=0)"),
#   runSimulation(A_sw,   R0=2, calib_mode="pre_removal_biology", T_R=3, TSPAN=TSPAN, n_runs=2,
#                 seed_mode="fixed", fixed_nodes=1:10, n_seeds=10, global_seed=8) |> mutate(kind="WS (small p)")
# )

# ggplot(df_sw, aes(time, prevalence, color = sim)) + geom_line() +
#   facet_wrap(~ kind, scales = "free_y") + theme_minimal() +
#   labs(title="Ring wavefront vs small-world shortcuts", x="Time", y="Prevalence")
```

### Reflection E

* Why can growth be close to linear on a ring lattice even when (R_0) is large?
* What do shortcuts change geometrically?

---

## Optional extension (choose ONE)

* **Seeding strategies:** compare `seed_mode="random"` vs `seed_mode="high_degree"` on PA.
* **Use your Prac1 contact network:** load an adjacency matrix you saved in Prac1 and compare it to PA/WS.

*(If you do this, include one figure + one paragraph.)*

---

## Write-up template (max 2 pages)

### A. Graph families + baseline epidemics (2–3 bullets)

* Bullet 1:
* Bullet 2:
* Bullet 3 (optional):

### B. Proxy vs early growth (short paragraph)

* What did `knbr-1` suggest?
* Did early growth agree?

### C. Calibration modes under removal (short paragraph)

* Which mode corresponds to “same biology”?
* What changes in interpretation?

### D. GC fraction and intervention (short paragraph)

* What does GC fraction capture?
* Why does targeting matter?

### E. Ring vs small-world (short paragraph)

* Describe the qualitative difference and why it happens.
