---
title: "MATH532 Practical 1 (Student) — Networks & Metrics (LO2)"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Setup

```{r libraries}
suppressPackageStartupMessages({
  library(igraph)
  library(tidyverse)
  library(MASS)
})

set.seed(532)

safe_mean_distance <- function(g) suppressWarnings(mean_distance(g, unconnected = TRUE))

giant_component_vertices <- function(g) {
  cc <- components(g)
  which(cc$membership == which.max(cc$csize))
}

TODO <- function(msg) stop(msg, call. = FALSE)
```

---

## Section A — Graph families & baseline metrics (guided)

### Instructions

1. Construct graphs:

   * Erdős–Rényi (G(N,p))
   * Ring lattice (Watts–Strogatz with (p=0))
   * Preferential attachment
   * Random modular graph
2. For each, construct a **degree-matched configuration model**.
3. Compute: mean(k), var(k), global clustering, average path length (APL), giant component fraction.
4. Present results in a table and write **2–3 bullet comparisons**.

### Familiarise

```{r A-familiarise}
g_demo <- erdos.renyi.game(200, p.or.m = 0.02, type = "gnp", directed = FALSE)

tibble(k = degree(g_demo)) |>
  ggplot(aes(k)) +
  geom_histogram(bins=40, color="white") +
  labs(title="Demo ER degree histogram") +
  theme_minimal()
```

### Provided builders

```{r A-builders}
build_er <- function(N, p = 0.0025) {
  erdos.renyi.game(N, p.or.m = p, type = "gnp", directed = FALSE)
}

build_ws <- function(N, nei = 5, p = 0) {
  g <- sample_smallworld(1, N, nei = nei, p = p)
  g <- as.undirected(g, "collapse")
  simplify(g)
}

build_pa <- function(N, m = 3) {
  sample_pa(N, m = m, directed = FALSE)
}

build_modular <- function(N, islands.n = 6, pin = 0.15, n.inter = 50) {
  g <- sample_islands(islands.n, floor(N / islands.n), pin, n.inter)
  simplify(g)
}

build_config_from_deg <- function(k) {
  g <- tryCatch(
    sample_degseq(k, method = "simple.no.multiple"),
    error = function(e) sample_degseq(k, method = "simple")
  )
  g <- as.undirected(g, "collapse")
  simplify(g)
}
```

### Task A: run graphs + compute metrics

```{r A-metrics}
N <- 3000
g_er   <- build_er(N)
g_ws0  <- build_ws(N, nei = 5, p = 0)
g_pa   <- build_pa(N, m = 3)
g_mod  <- build_modular(N)
g_conf <- build_config_from_deg(degree(g_er))

metrics_tbl <- tibble(
  model  = c("ER","WS0","PA","Modular","Config(deg=ER)"),
  mean_k = c(mean(degree(g_er)), mean(degree(g_ws0)),
             mean(degree(g_pa)), mean(degree(g_mod)), mean(degree(g_conf))),
  var_k  = c(var(degree(g_er)),  var(degree(g_ws0)),
             var(degree(g_pa)),  var(degree(g_mod)), var(degree(g_conf))),
  CC     = c(transitivity(g_er,"global"), transitivity(g_ws0,"global"),
             transitivity(g_pa,"global"), transitivity(g_mod,"global"),
             transitivity(g_conf,"global")),
  APL    = c(safe_mean_distance(g_er), safe_mean_distance(g_ws0),
             safe_mean_distance(g_pa), safe_mean_distance(g_mod),
             safe_mean_distance(g_conf)),
  GCfrac = c(max(components(g_er)$csize)/N,
             max(components(g_ws0)$csize)/N,
             max(components(g_pa)$csize)/N,
             max(components(g_mod)$csize)/N,
             max(components(g_conf)$csize)/N)
)

metrics_tbl
```

### Reflection A (write 2–3 bullets)

* Why does preferential attachment have high var(k)?
* Why is Config(deg=ER) triangle-poor despite matching degrees?

---

## Section B — CCDF (light coding)

### Instructions

1. Implement CCDF(k) = P(K ≥ k).
2. Plot CCDFs for ER, WS0, PA, Modular on log–log axes.
3. Why is CCDF preferable to histograms for diagnosing heavy tails?

### TODO function

```{r B-ccdf, eval=FALSE}
compute_ccdf <- function(k) {
  # TODO:
  # - tabulate k (table())
  # - turn into numeric tibble sorted by k
  # - pmf = Freq / sum(Freq)
  # - ccdf = reverse cumulative sum of pmf
  TODO("TODO (Section B): implement compute_ccdf(k)")
}
```

### Task B: plot CCDF

```{r B-plot, eval=FALSE}
ccdf_df <- bind_rows(
  compute_ccdf(degree(g_er))  |> mutate(model="ER"),
  compute_ccdf(degree(g_ws0)) |> mutate(model="WS0"),
  compute_ccdf(degree(g_pa))  |> mutate(model="PA"),
  compute_ccdf(degree(g_mod)) |> mutate(model="Modular")
)

ggplot(ccdf_df, aes(k, ccdf, color=model)) +
  geom_line() +
  scale_x_log10() + scale_y_log10() +
  theme_minimal() +
  labs(title="Degree CCDF", y="P(K ≥ k)")
```

### Reflection B

* Why does CCDF help detect heavy tails?

---

## Section C — Local clustering via (A^3) (Main task #1)

### Instructions

1. Triangle count per node: (T_i = (A^3)_{ii}/2).
2. Local clustering: (C_i = 2T_i / (k_i(k_i-1))) for (k_i \ge 2).
3. Global clustering: mean of valid (C_i).
4. Validate against `igraph::transitivity()`.

```{r C-stubs, eval=FALSE}
local_clustering_A3 <- function(A) {
  # INPUT: A is 0/1 symmetric adjacency matrix with zero diagonal
  # OUTPUT: numeric vector C_i
  # TODO:
  #   - k <- rowSums(A)
  #   - tri <- diag(A %*% A %*% A) / 2
  #   - compute C_i for k>=2
  TODO("TODO (Section C): implement local_clustering_A3(A)")
}

global_from_local <- function(A) {
  # TODO: mean of C_i over nodes with k>=2
  TODO("TODO (Section C): implement global_from_local(A)")
}
```

### Validation run

```{r C-validate, eval=FALSE}
idx <- sample(giant_component_vertices(g_ws0), 1200)
A_sub <- as.matrix(as_adj(g_ws0, sparse=FALSE))[idx, idx]
g_sub <- induced_subgraph(g_ws0, idx)

C_loc <- global_from_local(A_sub)
C_igr <- transitivity(g_sub, "global")

cat("Your global C (A^3) =", C_loc, "\n")
cat("igraph transitivity  =", C_igr, "\n")
```

### Reflection C

* Why does (\mathrm{diag}(A^3)/2) count triangles?

---

## Section D — Watts–Strogatz (C(p)/C(0)) and (L(p)/L(0)) (Main task #2)

### Instructions

1. Scan over (p \in {10^{-8},10^{-6},\dots,1}).
2. Compute and plot:

   * (C(p)/C(0))
   * (L(p)/L(0))
3. Identify “small-world” regime: (L) drops sharply while (C) remains high.

```{r D-stub, eval=FALSE}
ws_scan <- function(N, nei, p_grid, C0, L0) {
  # TODO:
  # - loop over p_grid
  # - build WS graph using build_ws()
  # - compute C = transitivity(...,"global")
  # - compute L = mean_distance(..., unconnected=TRUE)
  # - return tibble(p, C_ratio, L_ratio)
  TODO("TODO (Section D): implement ws_scan()")
}
```

```{r D-run, eval=FALSE}
p_grid <- c(1e-8,1e-6,1e-5,1e-4,1e-3,1e-2,1e-1,3e-1,1)
C0 <- transitivity(g_ws0,"global")
L0 <- safe_mean_distance(g_ws0)

sw <- ws_scan(N=3000, nei=5, p_grid=p_grid, C0=C0, L0=L0)

ggplot(sw, aes(p, C_ratio)) +
  geom_line() + scale_x_log10() +
  theme_minimal() + labs(title="C(p)/C(0)")

ggplot(sw, aes(p, L_ratio)) +
  geom_line() + scale_x_log10() +
  theme_minimal() + labs(title="L(p)/L(0)")
```

### Reflection D

* Mark the small-world regime and describe it in 2–3 sentences.

---

## Section E — Contact network from diary-like data (Main task #3)

### Instructions (pipeline)

1. Fit NegBin models for outside-household contacts by age band.
2. Create a synthetic population using:

   * household size distribution
   * **greedy household packing**
3. Sample outside-household contacts using fitted NegBin.
4. Build network:

   * household cliques
   * outside-household stub matching
5. Compute metrics and run a small sensitivity analysis.

**Greedy packing (definition):** assign households sequentially. Take next unassigned person (i), read their sampled household size (s_i), assign (i,\dots,i+s_i-1) to the same household, then continue.

> This section is intentionally more open-ended. If you get stuck, consult the answers script, but be able to explain each step.

### Load data (or use toy data)

```{r E-data}
file_path <- "contact_survey_clean.csv"
if (!file.exists(file_path)) {
  contacts_df <- tibble(
    id=1:2000,
    age=sample(0:80, 2000, replace=TRUE),
    hh_size=sample(1:6, 2000, replace=TRUE, prob=c(.30,.30,.18,.12,.06,.04)),
    contacts_outside_hh=rnbinom(2000, size=2, mu=6)
  )
} else {
  contacts_df <- readr::read_csv(file_path, show_col_types=FALSE)
}

cut_age <- function(x) cut(x, breaks=c(-Inf,12,25,65,Inf),
                           labels=c("child","young","adult","elder"),
                           right=FALSE)
```

### TODO functions

```{r E-stubs, eval=FALSE}
fit_nb_stratified <- function(df) {
  # INPUT: df with age, contacts_outside_hh
  # OUTPUT: tibble(band, size, mu)
  TODO("TODO (Section E): implement fit_nb_stratified(df)")
}

pack_households <- function(hh_sizes) {
  # INPUT: vector hh_sizes (length Nsyn)
  # OUTPUT: hh_id (length Nsyn) via greedy packing
  TODO("TODO (Section E): implement pack_households(hh_sizes)")
}

sample_outside_hh <- function(people, nb_table) {
  # INPUT: people with band; nb_table with band/size/mu
  # OUTPUT: vector contacts_outside_hh
  TODO("TODO (Section E): implement sample_outside_hh(people, nb_table)")
}

stub_match <- function(ids, counts, hh_id) {
  # INPUT: ids, counts, hh_id
  # OUTPUT: 2-col matrix of outside-household edges
  TODO("TODO (Section E): implement stub_match(ids, counts, hh_id)")
}

build_contact_network <- function(people) {
  # INPUT: people with id, hh_id, contacts_outside_hh
  # OUTPUT: igraph graph (HH cliques + outside HH edges; simplify)
  TODO("TODO (Section E): implement build_contact_network(people)")
}

sensitivity_sweep <- function(people, nb_table,
                              mu_mult = c(0.8,1,1.2),
                              size_mult = c(0.8,1,1.2)) {
  # OUTPUT: tibble(mu_mult, size_mult, mean_k, CC, APL, GCfrac)
  TODO("TODO (Section E): implement sensitivity_sweep(...)")
}
```

### Run pipeline

```{r E-run, eval=FALSE}
nb_by_band <- fit_nb_stratified(contacts_df)
nb_by_band

Nsyn <- 5000
hh_prob <- prop.table(table(contacts_df$hh_size))

people <- tibble(
  id = 1:Nsyn,
  hh_size = sample(as.integer(names(hh_prob)), Nsyn, replace=TRUE, prob=as.numeric(hh_prob))
)

people$hh_id <- pack_households(people$hh_size)
people$age <- sample(0:80, Nsyn, replace=TRUE)
people$band <- cut_age(people$age)

people$contacts_outside_hh <- sample_outside_hh(people, nb_by_band)

g_contact <- build_contact_network(people)

deg <- degree(g_contact)
metrics_contact <- tibble(
  mean_k = mean(deg),
  var_k  = var(deg),
  CC     = transitivity(g_contact, "global"),
  APL    = safe_mean_distance(g_contact),
  GCfrac = max(components(g_contact)$csize)/Nsyn
)
metrics_contact

sens_tbl <- sensitivity_sweep(people, nb_by_band)
sens_tbl
```

### Reflection E

* Which elements were estimated from data vs assumed?
* How do CC and APL change under ±20% NegBin parameter changes?

---

## Submission checklist (max 2 pages of narrative, plots allowed)

* A: metrics table + 2–3 bullets comparing models
* B: CCDF plot + 1–2 lines on CCDF vs histogram
* C: your (A^3) clustering code + validation output
* D: WS transition plots + interpretation
* E: NB fits, contact network metrics, sensitivity output + brief analysis
